Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-02-03 ms.qmd
First paragraph of Potential... finished. Start at second paragraph

### ***Potential evolutionary consequences for reduced plasticity and variance in physiologcial rates across ectotherms***

There is a general interest in understanding the interplay and relative importance of plasticity and genetic adaptation to a population's capacity to mitigate climate change effects [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate]. Phenotypic plasticity is expected to allow populations to persist in the face of climate extremes by better allowing phenotypes to match fitness optima, however, maladaptive responses can occur [@noble2019plastic; @radersma2020plasticity] and there may be limits / costs to plasticity that may prevent it from evolving [@Dewitt1998; but see @Murren2015]. We show that plastic adjustments in physiological rates and changes in their relative variance differs across habitats. In freshwater habitats, acclimation responses will likely be beneficial considering that these effects are larger than reductions in relative variance in physiological rates. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiolgical rates, suggesting acclimation may not be as beneficial. Relative variance reductions are particularly pronounced for terrestrial ectotherms and this coincides with a general inability to acclimate highlighting the vulnerability of terrestrial ectotherms to climate change. Considering acute responses of animals acclimated to high temperatures are generally slightly elevated compared to cold acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization is not likely to provide adaptive benefits under climate change. While it is unclear the extent to which such changes in variance are heritable, we expect changes in variance of this magnitude to impact the 'opportunity for selection' by reducing the strength of selection and/or impacting underlying genetic variance exposed to selection. Both are predicted to impact the capacity of populations to evolve under climate change [@hoffmann2011climate; @urban2023and].

>>>>

### ***Potential evolutionary consequences of reduced plasticity and variance in physiologcial rates across ectotherms***

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate]. Phenotypic plasticity can allow populations to persist in the face of climate extremes by matching phenotypes to fitness optima, although this is not always the case [@noble2019plastic; @radersma2020plasticity] and there may be limits and costs to plasticity that constrain its expression [@Dewitt1998; but see @Murren2015]. We show that plastic adjustments in physiological rates and changes in their relative variance differs across habitats. In freshwater habitats, acclimation responses will likely be beneficial considering that these effects are larger than reductions in relative variance in physiological rates. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiolgical rates, suggesting acclimation may not be as beneficial. Relative variance reductions are particularly pronounced in terrestrial ectotherms and this coincides with a general inability to acclimate. Considering acute responses of animals acclimated to high temperatures are generally slightly elevated compared to cold acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization is not likely to provide adaptive benefits under climate change. Together, these findings highlight the vulnerability of terrestrial ectotherms to climate change. It may be expected that reductions in variance of this magnitude impact the 'opportunity for selection' by reducing the strength of selection and/or impacting underlying genetic variance exposed to selection. Both are predicted to impact the capacity of populations to evolve under climate change [@hoffmann2011climate; @urban2023and].
